---
output: 
  github_document:
    toc: false
    fig_width: 7
    fig_height: 5
---
  
  <!-- README.md is generated from README.Rmd. Please edit that file -->
  
```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(CTAE)

options(digits = 2)

knitr::opts_chunk$set(
  collapse = TRUE,
  fig.path = "man/figures/",
  fig.width = 7,
  fig.height = 5,
  comment = "#>",
  dev = "png",
  dpi = 300
)
```
  
# CTAE - Canadian tree allometric equations
A collection of tools to calculate tree- or stand-level attributes developed for Canadian forests


## Models included so far

- Canadian national tree aboveground biomass equations (Lambert et al. 2005, Ung et al. 2008).

- Individual tree volume equations for major Alberta tree species (Huang 1994)

- Volume-to-biomass conversions models (Boudewyn et al. 2007)

- Total volume to merchantable volume conversions models (Boudewyn et al. 2007)


Updated model parameters for models developed by Boudewyn et al (2007) downloaded from https://nfi.nfis.org/en/biomass_models.



Boudewyn, P.A.; Song, X.; Magnussen, S.; Gillis, M.D. (2007). Model-based, volume-to-biomass conversion for forested and vegetated land in Canada. Natural Resources Canada, Canadian Forest Service, Pacific Forestry Centre, Victoria, BC. Information Report BC-X-411. 112 p.

Huang, S. (1994). Ecologically Based Individual Tree Volume Estimation for Major Alberta Tree Species. Report 1 - Individual tree volume estimation procedures for Alberta: Methods of Formulation and Statistical Foundations. Alberta Environmental Protection, Land and Forest Service, Forest Management Division, Edmonton, AB.

Lambert, M. C., Ung, C. H., & Raulier, F. (2005). Canadian national tree aboveground biomass equations. Canadian Journal of Forest Research, 35(8), 1996–2018. https://doi.org/10.1139/x05-112

Ung, C.-H., Bernier, P., & Guo, X.-J. (2008). Canadian national biomass equations: new parameter estimates that include British Columbia data. Canadian Journal of Forest Research, 38(5), 1123–1132. https://doi.org/10.1139/X07-224
 


## Installation 

You can install the most recent version of the package by executing the code below:

```{r, eval=FALSE}
devtools::install_github("ptompalski/CTAE")
library(CTAE)
```


# Additional details 



## Alberta Natural Subregion Codes
```{r, echo=FALSE}
CTAE::AlbertaNaturalRegSubreg %>%
  select(Code=NaturalSubregionCode, `Subregion name`=NaturalSubregion) %>%
  kable(format = "html")
```



## Huang et al. model availability (`V_Huang()`)

```{r, echo=FALSE}
parameters_HuangV %>% group_by(species, NaturalSubregionCode) %>% 
  count() %>%
  mutate(NaturalSubregionCode = if_else(is.na(NaturalSubregionCode), "Province", NaturalSubregionCode) ) %>%
  select(-n) %>%

  mutate(available="✅") %>%
  pivot_wider(names_from = NaturalSubregionCode, values_from = available) %>%
  relocate(Province,.after = species) %>%
  mutate(across(everything(), ~replace_na(.x, replace = ""))) %>%
  kable(format = "html")
```


<!-- ## Ecozone Codes -->

<!-- ```{r, echo=T} -->
<!-- CTAE::CodesEcozones -->

<!-- # CodesEcozones %>% -->
<!--   # select(`Ecozone code`=EcozoneCode, `Ecozone name` = EcozoneName) %>% -->
<!--   # kable(format = "html") -->
<!-- ``` -->