---
output: 
  github_document:
    toc: false
    fig_width: 7
    fig_height: 5
---
  
  <!-- README.md is generated from README.Rmd. Please edit that file -->
  
```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(CTAE)

options(digits = 2)

knitr::opts_chunk$set(
  collapse = TRUE,
  fig.path = "man/figures/",
  fig.width = 7,
  fig.height = 5,
  comment = "#>",
  dev = "png",
  dpi = 300
)
```
  
# CTAE: Canadian tree allometric equations

CTAE is an R package that brings together a collection of published allometric models developed for Canadian forests. The package provides a unified interface to estimate a range of tree- and stand-level attributes, based on models published in the Canadian forestry literature. The focus of CTAE is on standardization: models are implemented as faithfully as possible to their original formulations, with transparent parameter tables, consistent inputs/outputs, and jurisdiction-appropriate assumptions (e.g. merchantability rules).

CTAE currently includes allometric models to:

- Estimate tree-level volume (total and merchantable)
- Estimate aboveground biomass (AGB)
- Convert volume to biomass
- Convert total volume to merchantable volume
- Apply simple growth models to estimate changes in attributes over time



## Included models

### Aboveground biomass
- Canadian national tree aboveground biomass equations (Lambert et al. 2005, Ung et al. 2008).

### Volume (total and merchantable)


```{r results = "asis", echo = FALSE}
reg <- volume_model_registry_species()

# ---- collapse registry to one row per engine ----
reg_collapsed <- reg |>
  dplyr::group_by(engine) |>
  dplyr::summarise(
    scope = dplyr::first(scope),
    province_scope = list({
      x <- sort(unique(unlist(province_scope)))
      if (any(x == "ALL")) "ALL" else x
    }),

    # FIX: keep the actual description, not reference
    description = paste(unique(description), collapse = "; "),

    requires_ht_any = any(requires_ht),
    requires_ht_all = all(requires_ht),

    species = list(sort(unique(unlist(species)))),
    n_species = length(species[[1]]),

    reference = paste(unique(reference), collapse = "; "),
    .groups = "drop"
  )

# ---- helpers ----
fmt_geo <- function(scope, province_scope) {
  if (length(province_scope) == 1 && identical(province_scope, "ALL")) {
    "Canada (national)"
  } else {
    paste(province_scope, collapse = ", ")
  }
}

fmt_inputs <- function(any_ht, all_ht) {
  dplyr::case_when(
    !any_ht ~ "DBH",
    all_ht  ~ "DBH & height",
    TRUE    ~ "DBH or DBH & height"
  )
}

fmt_model_bullet_oneline <- function(
  engine,
  coverage,
  # inputs,
  n_species,
  description,
  reference
) {
  parts <- c(
    paste0("`", engine, "`"),
    paste0(description),
    paste0("Coverage: ", coverage),
    # paste0("Inputs: ", inputs),
    paste0(n_species, " species"),
    paste0("", reference)
  )

  paste0("- ", paste(parts, collapse = ". "))
}

# ---- build bullets ----
bullets <- reg_collapsed |>
  dplyr::mutate(
    Coverage = purrr::map2_chr(scope, province_scope, fmt_geo),
    Inputs = fmt_inputs(requires_ht_any, requires_ht_all)
  ) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    bullet = fmt_model_bullet_oneline(
      engine,
      Coverage,
      # Inputs,
      n_species,
      description,
      reference
    )
  ) |>
  dplyr::ungroup()

# ---- emit Markdown (important for README.Rmd) ----
cat(paste(bullets$bullet, collapse = "\n"))


```

#### `vol()`: automatic tree volume estimation across multiple models

`vol()` is a convenience wrapper for estimating total and merchantable tree volume using the volume models implemented in `CTAE`. Rather than requiring users to select a specific model, `vol()` consults an internal model registry and automatically determines which models are applicable for each tree based on:

- species availability in model parameter tables
- geographic scope (province / jurisdiction)
- required inputs (e.g. total height, subregion such as BEC zone)
- model ranking (regional models preferred over national where available)

Example:
```{r, echo=T, warning=F, message=F}
trees <- tibble::tibble(
  DBH = c(18, 22, 30, 26, 20),
  height = c(15, 18, 22, 20, 20),
  species = c("PICE.MAR", "BETU.PAP", "POPU.TRE", "PSEU.MEN", "PINU.BAN"),
  jurisdiction =  c("AB", "ON", "QC", "BC", "MB"),
  subregion = c(NA, NA, NA, "CWH", "Boreal Plains")
)

trees |>
  dplyr::mutate(
    vol(
      DBH = DBH,
      height = height,
      species = species,
      jurisdiction = jurisdiction,
      subregion = subregion,
      keep_model_id = TRUE
    )
  )
```


#### Merchantability criteria
Merchantable volume in CTAE is defined using jurisdiction-specific merchantability rules, following the officially adopted criteria in each province or territory (e.g., minimum top diameter, minimum DBH, and stump height). For most models, these rules are applied dynamically based on the provided jurisdiction.

Some volume models, however, have merchantability criteria fixed within the model formulation itself, reflecting how the original equations were developed and calibrated. In these cases (e.g., `vol_honer83`, `vol_sharma2021`, `vol_fortin2007`, `vol_nigh2016`), merchantable volume is computed using the model-specific, hard-coded criteria and does not vary by jurisdiction or species.

The jurisdiction-specific merchantability criteria used by CTAE can be inspected in the internal dataset `merchcrit`, which documents the values applied for each jurisdiction (and, where applicable, by species or subregion).

### Other

- National volume-to-biomass conversions models (Boudewyn et al. 2007)
- National total volume to merchantable volume conversions models (Boudewyn et al. 2007)
- National growth and yield model (H, BA, V) (Ung et al. 2009)


## Installation 

You can install the most recent version of the package by executing the code below:

```{r, eval=FALSE}
devtools::install_github("ptompalski/CTAE")
library(CTAE)
```

## License

This package is licensed under the GNU Lesser General Public License (LGPL-3.0).

© His Majesty the King in Right of Canada, as represented by the Minister of Natural Resources, 2026.

See the `LICENSE` file for full license terms.

## References 

Boudewyn, P.A.; Song, X.; Magnussen, S.; Gillis, M.D. (2007). Model-based, volume-to-biomass conversion for forested and vegetated land in Canada. Natural Resources Canada, Canadian Forest Service, Pacific Forestry Centre, Victoria, BC. Information Report BC-X-411. 112 p.

Fortin, M., DeBlois, J., Bernier, S., Blais, G., 2007. Mise au point d’un tarif de cubage général pour les forêts québécoises : une approche pour mieux évaluer l’incertitude associée aux prévisions. The Forestry Chronicle 83, 754–765. https://doi.org/10.5558/tfc83754-5


Gal, J., & Bella, I.E. (1994). New stem taper functions for 12 Saskatchewan timber species. Natural Resources Canada, Canadian Forest Service, Northwest Region, Information Report NOR-X-338. Table 5.

Honer, T.G.; Ker, M.F.; Alemdag, I.S. 1983. Metric timber tables for the commercial tree species of central and eastern Canada. Environ. Can., Can. For. Serv., Maritimes For. Res. Cent., Fredericton, NB. Inf. Rep. M-X-140. https://ostrnrcan-dostrncan.canada.ca/handle/1845/239814

Huang, S. (1994). Ecologically Based Individual Tree Volume Estimation for Major Alberta Tree Species. Report 1 - Individual tree volume estimation procedures for Alberta: Methods of Formulation and Statistical Foundations. Alberta Environmental Protection, Land and Forest Service, Forest Management Division, Edmonton, AB.

Kozak, A. (1988). A variable-exponent taper equation. Canadian Journal of Forest Research, 18, 1363–1368

Kozak, A. (1994). Development of Taper Equations by BEC Zones and Species. Province of British Columbia, Ministry of Forests (report).

Klos, R. J., Wang, G. G., Dang, Q.-L., & East, E. W. (2007). Taper equations for five major commercial tree species in Manitoba, Canada. Western Journal of Applied Forestry, 22(3), 163–170.

Lambert, M. C., Ung, C. H., & Raulier, F. (2005). Canadian national tree aboveground biomass equations. Canadian Journal of Forest Research, 35(8), 1996–2018. https://doi.org/10.1139/x05-112

Nigh, G.D., 2016. Total and merchantable volume equations for common tree species in British Columbia: by region and biogeoclimactic zone (No. Prov. B.C., Victoria, B.C. Tech. Rep. 106.).

Sharma, M. (2021). Total and Merchantable Volume Equations for 25 Commercial Tree Species Grown in Canada and the Northeastern United States. Forests, 12, 1270.

Ung, C.H., Bernier, P., & Guo, X.-J. (2008). Canadian national biomass equations: new parameter estimates that include British Columbia data. Canadian Journal of Forest Research, 38(5), 1123–1132. https://doi.org/10.1139/X07-224
 
Ung, C.H., Bernier, P.Y., Guo, X.J., Lambert, M.-C., 2009. A simple growth and yield model for assessing changes in standing volume across Canada’s forests. The Forestry Chronicle 85, 57–64. https://doi.org/10.5558/tfc85057-1

Ung, C.H., Guo, X.J., Fortin, M., 2013. Canadian national taper models. Forestry Chronicle 89, 211–224. https://doi.org/10.5558/tfc2013-040

Zakrzewski, W.T., Penner, M., 2013. A comparison of tree stem taper models for use in Ontario. Ontario Forest Research Institute, Report 176.